---
title: "Interpolating Bathymetry and Estimating Habitat"
output:
  pdf_document: default
  html_notebook: default
---

In this R notebook, we will calculate the percentage of adult bottomfish habitat (100-400m) in each of Hawaii's statistical fishery reporting areas, and calculate the fraction of that habitat that is protected by the Bottomfish Restricted Fishing Area (BRFA) network. 

However, the 50m bathymetry we will use to do this is missing values for some key areas. We will first need to gap fill the missing bathymetry values. 

### Setup
First we set up our workspace enviornment including loading all required packages, our data files, and setting a timer for our script
```{r}
## Starting Script Timer
sTimer = proc.time()

## Establishing working directory
proj_dir = "/Volumes/GoogleDrive/My Drive/Weng Lab/Personal_Folders/Steve/dissertation work/Ch 5. BACIP Analysis"

## Loading Packages
library('marmap')
library('doParallel')
library('beepr')
library('spatial.tools')
library('maptools')
library('rgdal')
library('raster')

#### Loading Spatial Data
bty_50 = raster(file.path(proj_dir, 'data/himbsyn.bathy.v19.grd')) #bathymetry at 60m resolution
bty_1 = raster(file.path(proj_dir, 'data/himbsyn.bathytopo.1km.v19.grd')) #bathymetry at 1km resolution
coast = readOGR(file.path(proj_dir, 'data/coast/coast_n83.shp')) #coastline

brfa08 = readOGR(file.path(proj_dir, 'data/brfa07/BRFA_Closed.shp')) #current BRFA placements
brfa99 = readShapePoly(file.path(proj_dir, 'data/brfa98/MHIRFAs.shp')) #old BRFA placements
stat_grids = readOGR(file.path(proj_dir, 'data/mhi_brfa_fishgrid_map.mpk Folder/commondata/regs/Fishchart2008.shp'))
```

### Cleaning spatial data
Now we need to get our spatial data into a similar projection. 

We'll adjust the crs of our bathymetry sources and plot them along with the statistical grids to ensure they properly overlap. 

Finally, We'll also mask off our raster data so we're not doing interpolation over the entire map (which is very large) but instead over a subset that we're interested in.
```{r}
### Adjusting bathymetry CRS
crs(bty_1)<-crs(bty_50)

### Plotting bathymetry data overlayed with statistical grids 
plot(bty_50)
plot(stat_grids, border = 'black', add = T)

plot(bty_1)
plot(stat_grids, border = 'black', add = T)

#### Modifying the extent of our raster objects
### Creating a new extent object for the bbox_to_SpatialPolygons function
ext_obj = extent(bty_50)
ext_obj@xmin = -160 
ext_obj@xmax = -154.13
ext_obj@ymin = 18.33 
ext_obj@ymax = 23

### Masking off our raster file
 mhi_bbox = bbox_to_SpatialPolygons(extent(bty_1), proj4string = crs(bty_50))
 bty_1 = mask(x = bty_1, mask = mhi_bbox)
 bty_50 = mask(x = bty_50, mask = mhi_bbox)
```


### Interpolating Missing Raster Data
Now that our data is alligned, we'll use the 1km bathymetry to fill in missing values in the 50m bathymetry. We do this by first resampling the 1km set using the 50m raster as a template. This will chop the 1km set up into 50m subsections, though each 20x20 pixel region will have the same value as the overlayed 1km data. 

Then we will fill in our missing 50m values with the values from the resampled 1km data
```{r}
## Resample 1 km to 50 m
 bty_1_resample = raster::resample(x = bty_1, y = bty_50, method = 'ngb')

 na_ind_50 = is.na(getValues(bty_50))
 bty_50_2 = bty_50
 bty_50_2[na_ind_50] = bty_1_resample[na_ind_50]
 
```


We've filled in our missing NA values, but as mentioned, the 20x20 pixel regions we've filled are uniform in depth because they were smapled from the 1km grid.

We can smooth out all of this using bilinear interpolation. This will produce a smoothed continuous bathymetry surface that models  depth across those missing pixel regions. 

This is done by resampling our combined data set using a blank raster template of identical dimensions.

After we have smoothed our data, we'll replace the NA values missing from the original 50m data with the smoothed interpolated values while we retain the known values values from the original data. We'll save our final raster out so we never have to do this again.
```{r}
### Make new raster from interpolated 50m set with only interpolated points from 1km resample raster
 bty_50_empty = bty_50
 bty_50_empty[!is.na(bty_50_empty)] = NA
 
### Using bilinear interpolation to resample the entire composited map (50m gap filled with 1km)
  bty_50_resample_all = raster::resample(x = bty_50_2, bty_50_empty, method = "bilinear")
  
## Now combine resampled and interpolated 50m back with 50m data we know is good
  bty_50_composited_by_bty_50_resample = bty_50
  bty_50_composited_by_bty_50_resample[na_ind_50] = bty_50_resample_all[na_ind_50]
  
### Plotting our composited bathymetry raster
  plot(bty_50_composited_by_bty_50_resample)
  
### Save our composited raster data
writeRaster(bty_50_composited_by_bty_50_resample, file.path(proj_dir, '50m Bathy Resampled from Mosaic'))
bty = bty_50_composited_by_bty_50_resample
```

We've now successfully gap filled our missing bathymetry!

### Calculating Habitat in Each Reporting Area
Now we will calculate the habitat (100 - 400m) in each statistical reporting grid as well as the amount of habitat that is protected under each BRFA management regime.

We'll begin by getting our data into the same CRS object, removing offshore statistical grids we know do not have any bottomfish habitat, and then plotting everything on top of one another to make sure our data remains aligned.
```{r}
### Reproject stat_grids
bty = projectRaster(bty, crs = '+proj=utm +zone=4 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0 ')
stat_grids = spTransform(stat_grids, CRSobj = crs(bty))

## Remove statistical reporting grids we don't care about
reporting_grids_to_keep = c(100:128, 300:333, 351, 400:429, 452,500:528)
stat_grids = stat_grids[stat_grids$AREA_ID %in% reporting_grids_to_keep, ]

## Plot everything to make sure it looks right
pdf('Plot of Reporting Grids, BRFAs, and Bathy.pdf', width = 11, height = 8)
plot(bty)
plot(stat_grids, border = 'black', add = T)
plot(coast, col= 'tan', add = T)
plot(brfa99, lwd = 2, add = T, border = 'red')
plot(brfa08, add = T, lwd = 2, border = 'purple')
dev.off()
```


Next we loop through each statistical area to calculate the amount of total and protected habitat contained within.

We start by defining our depth strata in 50m increments. Then we'll use the union function to subdivide our statistical grid and BRFA boundary shapefiles. This will create a single object with subdivided reporting areas for protected and non-protected habitat
```{r}
## Defining our the depth cuts that bound each of our depth strata
depth_cuts = seq(0, -500, by = -50)

#### 1999 
## Subdivide our stats grid to include BRFA Boundaries
stat_grids_by_brfa99 = union(stat_grids, brfa99)
## NA names throw an error so we'll just call them "NA" so we can find them later
stat_grids_by_brfa99$AREA_ID[is.na(stat_grids_by_brfa99$AREA_ID)] = 'NA'
```

Now we'll loop through each repoting grid to calculate the total and protected habitat contained within. We we set up dataframe for our results and then extract all bathymetry values under each reporting grid shapefile. 

The shapefiles contain the amount of squared area contained within so to calculate the area, we multiply the total area by the percentage of cells matching a given depth strata. 

For example, if an reporting area is 10km in size and 20% of cell values are within 50 and 100m, then we can say that area contains 5km of habitat between 50 and 100m. We then do the same thing using only the shapefiles that are "protected" as defined by the ID column.

We'll also divide the protected habitat area by total habitat area to get the % protected habitat.

For each reporting area, we'll write the amount of habitat back to our results data frame and then save our results for future analysis.
```{r}
## Creating output dataframe
reporting_areas_1999 = data.frame(stringsAsFactors = FALSE)

## Looping through each reporting area
for(i in 1:length(unique(stat_grids$AREA_ID))){ 
  print(paste(i, 'of', length(unique(stat_grids$AREA_ID))))
  area_data = c()
  area_id = unique(stat_grids$AREA_ID)[i]
  
  ## Calculating total area for the entire reporting grid
  total_grid_area = sum(area(stat_grids[stat_grids$AREA_ID == area_id, ])) *10^-6 # in km2
  ## Getting all data associated with this grid
  all_grid_data = unlist(extract(bty, stat_grids_by_brfa99[stat_grids_by_brfa99$AREA_ID == area_id, ]))
  
  ### Calculating total habitat area in slices
  habitat_area_by_depth = as.data.frame(matrix(data = 0, nrow = 1, ncol = length(depth_cuts)-1))
  habitat_area_colnames = c()
  for(j in 2:length(depth_cuts)){
    ## Get the area within each depth slice by calculating the number of cells meeting depth requirements, then dividing by the number of total cells to get % in depth range, then multiply by total area
    habitat_area_by_depth[j-1] = length(all_grid_data[all_grid_data <= depth_cuts[j-1] & all_grid_data > depth_cuts[j] & !is.na(all_grid_data)]) / length(all_grid_data) * total_grid_area
    habitat_area_colnames = c(habitat_area_colnames, paste('habitat_area_total', abs(depth_cuts[j-1]), abs(depth_cuts[j]), sep = "_"))
  }
  ## Assigning generated column names
  colnames(habitat_area_by_depth) = habitat_area_colnames
  
  ### Calculating BRFA habitat area in slices
  brfa_habitat_area_by_depth = as.data.frame(matrix(data = 0, nrow = 1, ncol = length(depth_cuts)-1))
  brfa_habitat_area_colnames = c()
  ## If a reporting grid has a BRFA
  if(dim(stat_grids_by_brfa99[stat_grids_by_brfa99$AREA_ID == area_id & !is.na(stat_grids_by_brfa99$ID), ])[1] > 0){
    ## Loop through each depth strata and calculate the number of cells within.
    for(j in 2:length(depth_cuts)){
      brfa_bathy_data = unlist(extract(bty, stat_grids_by_brfa99[stat_grids_by_brfa99$AREA_ID == area_id & !is.na(stat_grids_by_brfa99$ID), ]))
      ## Calculating total area of each strata by computing the % of total cells falling within that strata and multiplying it by total calculated
      brfa_habitat_area_by_depth[j-1] = length(brfa_bathy_data[brfa_bathy_data <= depth_cuts[j-1] & brfa_bathy_data > depth_cuts[j] & !is.na(brfa_bathy_data)]) / length(all_grid_data) * total_grid_area
      ## get our column names that define each strata
      brfa_habitat_area_colnames = c(brfa_habitat_area_colnames, paste('habitat_area_brfa', abs(depth_cuts[j-1]), abs(depth_cuts[j]), sep = "_"))
    }
  } else {
    ## We still need to get our column names that define each strata
    for(j in 2:length(depth_cuts)){
      brfa_habitat_area_colnames = c(brfa_habitat_area_colnames, paste('habitat_area_brfa', abs(depth_cuts[j-1]), abs(depth_cuts[j]), sep = "_"))
    }
  }
  ## Assigning generated column names
  colnames(brfa_habitat_area_by_depth) = brfa_habitat_area_colnames
  
  ## Writing out data frame
  reporting_areas_1999 = rbind(reporting_areas_1999, data.frame('grid' = area_id, 'habitat_area_total' = sum(habitat_area_by_depth), habitat_area_by_depth, 'habitat_area_brfa_total' = sum(brfa_habitat_area_by_depth), brfa_habitat_area_by_depth))
}

## getting percent of adult habitat (100-400m) in each reporting area that's protected 
reporting_areas_1999$percent_protected_1999 = rowSums(reporting_areas_1999[ ,which(grepl('habitat_area_brfa_', colnames(reporting_areas_1999)) & !grepl('brfa_0', colnames(reporting_areas_1999)) & !grepl('brfa_500', colnames(reporting_areas_1999)) & !grepl('habitat_area_brfa_total', colnames(reporting_areas_1999)) & !grepl('brfa_450', colnames(reporting_areas_1999)))]) / 
  rowSums(reporting_areas_1999[ ,which(grepl('habitat_area_total_', colnames(reporting_areas_1999)) & !grepl('total_0', colnames(reporting_areas_1999)) & !grepl('total_500', colnames(reporting_areas_1999)) & !grepl('total_450', colnames(reporting_areas_1999)))])
reporting_areas_1999$percent_protected_1999[is.nan(reporting_areas_1999$percent_protected_1999)] = 0
## Writing out file if percent protected ranges between 0 and 1 and no BRFA habitat is greater than total habitat

if(max(range(reporting_areas_1999$percent_protected_1999)) <= 1 & min(range(reporting_areas_1999$percent_protected_1999)) >= 0 & all(reporting_areas_1999$habitat_area_total >= reporting_areas_1999$habitat_area_brfa_total)) {
  ## Reordering our reporting areas by area id (grid)
  reporting_areas_1999 = reporting_areas_1999[order(as.numeric(reporting_areas_1999$grid)), ]
  ## Writing out our dataframe as a .csv file
  write.csv(reporting_areas_1999, file.path(proj_dir, "data/reporting_grid_1999_area_data.csv"), row.names = F)
} else {
  print('Error with 1999 reporting areas')
  beep(1)
}
```


Now we repeat this same process for the 2008 revised BRFA network.
```{r}
## Subdivide our stats grid to include BRFA Boundaries
stat_grids_by_brfa08 = union(stat_grids, brfa08)
## NA names throw an error so we'll just call them "NA" so we can find them later
stat_grids_by_brfa08$AREA_ID[is.na(stat_grids_by_brfa08$AREA_ID)] = 'NA'

## Looping through each reporting grid
reporting_areas_2008 = data.frame(stringsAsFactors = F)
# foreach(i = 1:length(unique(stat_grids$AREA_ID)), .combine = rbind) %dopar% {
for(i in 1:length(unique(stat_grids$AREA_ID))){ ; print(paste(i, 'of', length(unique(stat_grids$AREA_ID))))
  area_data = c()
  area_id = unique(stat_grids$AREA_ID)[i]
  
  ## Calculating total area for the entire reporting grid
  total_grid_area = sum(area(stat_grids[stat_grids$AREA_ID == area_id, ])) *10^-6 # in km2
  ## Getting all data associated with this grid
  all_grid_data = unlist(extract(bty, stat_grids_by_brfa08[stat_grids_by_brfa08$AREA_ID == area_id, ]))
  
  ### Calculating total habitat area in slices
  habitat_area_by_depth = as.data.frame(matrix(data = 0, nrow = 1, ncol = length(depth_cuts)-1))
  habitat_area_colnames = c()
  for(j in 2:length(depth_cuts)){
    ## Get the area within each depth slice by calculating the number of cells meeting depth requirements, then dividing by the number of total cells to get % in depth range, then multiply by total area
    habitat_area_by_depth[j-1] = length(all_grid_data[all_grid_data <= depth_cuts[j-1] & all_grid_data > depth_cuts[j] & !is.na(all_grid_data)]) / length(all_grid_data) * total_grid_area
    habitat_area_colnames = c(habitat_area_colnames, paste('habitat_area_total', abs(depth_cuts[j-1]), abs(depth_cuts[j]), sep = "_"))
  }
  colnames(habitat_area_by_depth) = habitat_area_colnames
  
  ### Calculating BRFA habitat area in slices
  brfa_habitat_area_by_depth = as.data.frame(matrix(data = 0, nrow = 1, ncol = length(depth_cuts)-1))
  brfa_habitat_area_colnames = c()
  ## If a reporting grid has a BRFA
  if(dim(stat_grids_by_brfa08[stat_grids_by_brfa08$AREA_ID == area_id & !is.na(stat_grids_by_brfa08$BRFA_Let), ])[1] > 0){
    ## Loop through each depth strata and calculate the number of cells within.
    for(j in 2:length(depth_cuts)){
      brfa_bathy_data = unlist(extract(bty, stat_grids_by_brfa08[stat_grids_by_brfa08$AREA_ID == area_id & !is.na(stat_grids_by_brfa08$BRFA_Let), ]))
      ## Calculating total area of each strata by computing the % of total cells falling within that strata and multiplying it by total calculated
      brfa_habitat_area_by_depth[j-1] = length(brfa_bathy_data[brfa_bathy_data <= depth_cuts[j-1] & brfa_bathy_data > depth_cuts[j] & !is.na(brfa_bathy_data)]) / length(all_grid_data) * total_grid_area
      ## get our column names that define each strata
      brfa_habitat_area_colnames = c(brfa_habitat_area_colnames, paste('habitat_area_brfa', abs(depth_cuts[j-1]), abs(depth_cuts[j]), sep = "_"))
    }
  } else {
    ## We still need to get our column names that define each strata
    for(j in 2:length(depth_cuts)){
      brfa_habitat_area_colnames = c(brfa_habitat_area_colnames, paste('habitat_area_brfa', abs(depth_cuts[j-1]), abs(depth_cuts[j]), sep = "_"))
    }
  }
  colnames(brfa_habitat_area_by_depth) = brfa_habitat_area_colnames
  
  #return(data.frame('grid' = area_id, 'habitat_area_total' = sum(habitat_area_by_depth), habitat_area_by_depth, 'habitat_area_brfa_total' = sum(brfa_habitat_area_by_depth), brfa_habitat_area_by_depth))
  reporting_areas_2008 = rbind(reporting_areas_2008, data.frame('grid' = area_id, 'habitat_area_total' = sum(habitat_area_by_depth), habitat_area_by_depth, 'habitat_area_brfa_total' = sum(brfa_habitat_area_by_depth), brfa_habitat_area_by_depth))
}

## getting percent of adult habitat (100-400m) in each reporting area that's protected 
reporting_areas_2008$percent_protected_2008 = rowSums(reporting_areas_2008[ ,which(grepl('habitat_area_brfa_', colnames(reporting_areas_2008)) & !grepl('brfa_0', colnames(reporting_areas_2008)) & !grepl('brfa_500', colnames(reporting_areas_2008)) & !grepl('habitat_area_brfa_total', colnames(reporting_areas_2008)) & !grepl('brfa_450', colnames(reporting_areas_2008)))]) / 
  rowSums(reporting_areas_2008[ ,which(grepl('habitat_area_total_', colnames(reporting_areas_2008)) & !grepl('total_0', colnames(reporting_areas_2008)) & !grepl('total_500', colnames(reporting_areas_2008)) & !grepl('total_450', colnames(reporting_areas_2008)))])
reporting_areas_2008$percent_protected_2008[is.nan(reporting_areas_2008$percent_protected_2008)] = 0
## Reordering our reporting areas by area id (grid)
reporting_areas_2008 = reporting_areas_2008[order(as.numeric(reporting_areas_2008$grid)), ]

## Writing out file if percent protected ranges between 0 and 1 and no BRFA habitat is greater than total habitat
if(max(range(reporting_areas_2008$percent_protected_2008)) <= 1 & min(range(reporting_areas_2008$percent_protected_2008)) >= 0 & all(reporting_areas_2008$habitat_area_total >= reporting_areas_2008$habitat_area_brfa_total)) {
  ## Writing out our dataframe as a .csv file
  write.csv(reporting_areas_2008, file.path(proj_dir, "data/reporting_grid_2008_area_data.csv"), row.names = F)
} else {
  print('Error with 2008 reporting areas')
  beep(1)
}
```

### Cleanup
And we're done! In this final chunk we'll calculate how long it took our script to run and send a notification to my phone so that I know everything ran successfully. 
```{r}

run_time = (proc.time()[3] - sTimer[3]) / 60
send_push(user = 'uGEHvA4hr37tsrCCtpSv4sUUxVuTqN', message = paste('Habitat area by stat grids and BRFAs calculated in:', run_time, ' minutes', sep = ""))
```